---
title: "Multi-omics data integration: MOMS-PI example"
author: "Margaret Turner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-omics data integration: MOMS-PI example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(tidyverse)
library(magrittr)
library(phyloseq)
library(MultiAssayExperiment)
library(knitr)
library(kableExtra)

library(TidyMultitude)
```

This vignette reproduces the **Multi-omics data integration: MOMS-PI example** section from the [Working with open-source Human Microbiome Project Data phases 1 (HMP) and 2 (iHMP): Efficient Data Access and Analysis Workflow](http://biocworkshops2019.bioconductor.org.s3-website-us-east-1.amazonaws.com/page/MicrobiomeWorkshop__MicrobiomeWorkshop/) workshop, demonstrating how the pre-processing steps can be simplified using `TidyMultitude`.

First, we will walk through the steps of the workflow used in the HMP workshop. Then, we will demonstrate how these steps can be simplified using `TidyMultitude`.

## Setup

Both workflows will use data from the [`HMP2Data package`](https://bioconductor.org/packages/release/data/experiment/html/HMP2Data.html).

```{r}
# Load 16S rRNA data
data("momspi16S_mtx")
data("momspi16S_samp")
data("momspi16S_tax")

# Load cytokines data
data("momspiCyto_mtx")
data("momspiCyto_samp")
```

### Objects for HMP workshop workflow

For the cytokine data, the co-inertia analysis only relies on the sample information matrix `momspiCyto_samp` and the cytokines matrix `momspiCyto_mtx`, so those will be left as is.

The 16S rRNA data can be contained in a `phyloseq` object.

```{r}
# Construct phyloseq object
momspi16S_phyloseq <- phyloseq(otu_table(momspi16S_mtx, taxa_are_rows = TRUE),
                               sample_data(momspi16S_samp),
                               tax_table(momspi16S_tax))
momspi16S_phyloseq
```


### Objects for TidyMultitude workflow

For the TidyMultitude workflow, we will need a `MultiAssayExperiment` object containing `SummarizedExperiment` objects representing the cytokine and 16S rRNA data, respectively. `MultiAssayExperiment` objects cannot contain `phyloseq` objects, so we must make a `SummarizedExperiment` for our 16S rRNA data.

```{r}
# Construct cytokines SummarizedExperiment
momspiCyto <- SummarizedExperiment(assays = list(cyto_conc = momspiCyto_mtx),
                                   colData = momspiCyto_samp,
                                   rowData = data.frame(cytokine = 
                                                        rownames(momspiCyto_mtx)))

# Construct 16S SummarizedExperiment
momspi16S <- SummarizedExperiment(assays = list(counts = momspi16S_mtx),
                                  colData = momspi16S_samp,
                                  rowData = momspi16S_tax)

# Construct MultiAssayExperiment
momspi_data <- MultiAssayExperiment(experiments = list(phy16S = momspi16S,
                                                       cyto = momspiCyto))

momspi_data
```

## HMP workshop workflow

```{r}
# Combine 16S and cytokines data

# order both sets by visit number within a subject
momspi16S_samp <- momspi16S_samp[
  with(momspi16S_samp, order(subject_id, sample_body_site, visit_number)),
] 

momspiCyto_samp <- momspiCyto_samp[
  with(momspiCyto_samp, order(subject_id, sample_body_site, visit_number)),
]

# Select data collected at the same visit
combined_samp <- merge(momspi16S_samp, momspiCyto_samp, 
                       by = c("subject_id", "sample_body_site", 
                        "project_name", "study_full_name",
                        "subject_gender", "subject_race",
                        "visit_number"))

# Select data from first visit only
combined_samp <- combined_samp[combined_samp$visit_number ==  1,]

#select 16S data for those samples
combined_16S_phyloseq <- subset_samples(momspi16S_phyloseq, file_name %in% combined_samp$file_name.x)

# get rid of otus that are not observed in any sample for this subset
combined_16S_phyloseq %<>%
    taxa_sums() %>%
    is_greater_than(0) %>%
    prune_taxa(combined_16S_phyloseq)
combined_16S_mtx <- otu_table(combined_16S_phyloseq)


combined_Cyto_mtx <- momspiCyto_mtx[, colnames(momspiCyto_mtx) %in% combined_samp$file_name.y ]

# Make sure that samples are in rows and variables (taxa and cytokines) are in columns.
combined_16S_mtx <- t(combined_16S_mtx)
combined_Cyto_mtx <- t(combined_Cyto_mtx)

#Taxa are converted to proportions.
combined_16S_mtx <- combined_16S_mtx/apply(combined_16S_mtx, 1, sum)

# Center 16S data to perform PCA on the covariance matrix
taxa_mtx <- scale(combined_16S_mtx, center = TRUE, scale = FALSE)

# Center and scale cytokines data to perform PCA on the correlation matrix
cyto_mtx <- scale(combined_Cyto_mtx, center = TRUE, scale = TRUE)

# Normalize magnitude and perform PCA
taxa_tr <- sum(taxa_mtx*taxa_mtx)/(dim(taxa_mtx)[1]-1)
taxa_mtx <- taxa_mtx/sqrt(taxa_tr)
taxa.pca <- ade4::dudi.pca(taxa_mtx, scannf=FALSE, nf =61,
                     center = FALSE, scale = FALSE)

cyto_tr <- sum(cyto_mtx*cyto_mtx)/(dim(cyto_mtx)[1]-1)
cyto_mtx <- cyto_mtx/sqrt(cyto_tr)
cyto.pca <- ade4::dudi.pca(cyto_mtx, scannf=FALSE, nf =61,
                     center = FALSE, scale = FALSE)

# Co-inertia is available through R package ade4; it takes ade4 PCA objects and performs joint eigendecomposition.
coin_hmp <- ade4::coinertia(taxa.pca, cyto.pca, scannf = FALSE, nf = 2)

# RV coefficient – measure of similarity between 16S and cytokines tables
coin_hmp$RV
#> [1] 0.0396224
```

### Results

```{r}
kable(head(coin_hmp$co),
      digits = 5)%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r}
kable(head(coin_hmp$li),
      digits = 5)%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r, echo=FALSE}
# PlotCoinVars function from workshop -- modified to include a graph title
PlotCoinVars <- function(coin, title,
                         tab1 = "Table1", tab2 = "Table2", 
                         Labels1 = NULL, Labels2 = NULL, label = TRUE,
                         hjust = 0, vjust = -1.5, PtSize = 2, LblSize = 2){

  if(is.null(Labels1)){Labels1 <- 1:nrow(coin$co)} #tab1 labels
  if(is.null(Labels2)){Labels2 <- 1:nrow(coin$li)} #tab2 labels
  
  #extract scores for each table
  x = colnames(coin$co)[1]
  y = colnames(coin$co)[2]
  
  #first table data
  df1 <- data.frame(coin$co$Comp1, coin$co$Comp2, Labels1, rep(tab1, nrow(coin$co)))
  rownames(df1) <- rownames(coin$co)
  names(df1) <- c(x,y,  "Labels", "Table")
  
  #second table data
  df2 <- data.frame(coin$li$Axis1, coin$li$Axis2, Labels2, rep(tab2, nrow(coin$li)))
  rownames(df2) <- rownames(coin$li)
  names(df2) <- c(x,y, "Labels", "Table")
  
  #conbine two tables for plotting
  df <- rbind(df1, df2)
  ord_map = aes_string(x = x, y = y, color = "Table", shape = "Table")
  
  CW_X <- ggplot(df, ord_map) + geom_point( size = PtSize) + 
          scale_color_manual(values = c("red", "blue")) + theme_bw() +
          labs(title = title, x = "", y = "") + 
          theme(legend.title=element_blank())
  
  if(label == TRUE) { 
      #Last thing: Fix order of labels
      lbl_map = aes_string(x = x, y = y, label = "Labels")  
      CW_X <- CW_X + geom_text(data = df, mapping = lbl_map, size = LblSize, vjust = vjust, hjust = hjust)
  }
  return(CW_X)
}

plot1 <- PlotCoinVars(coin_hmp, title = "HMP Workshop workflow",
             tab1 = "taxa", tab2 = "cytokines", 
             Labels1 = NULL,
             Labels2 = rownames(coin_hmp$li),
             label = TRUE,PtSize=2,LblSize=3,
             hjust = 0, vjust = -1.5)
plot1
```

## TidyMultitude workflow

First, we will create a function that we can use later in `transmute()`. This function centers (and optionally scales) a transposed matrix then performs the same total-variance normalization used in the HMP workshop.

```{r}
# Define a preprocessing function
prep <- function(data, scale = c(TRUE, FALSE)){
  # Note that the data is transposed so that these steps are performed rowwise
  scaled <- scale(t(data), center = TRUE, scale = scale) # Center and scale
  tr <- sum(scaled * scaled)/(dim(scaled)[1]-1) # Fast trace computation
  processed <- scaled/sqrt(tr) # Normalize magnitude
  return(t(processed)) # Rotate back
}
```



```{r}
pca_list <- momspi_data %>% 
  # Order both sets by visit number within a subject
  arrange_colData(subject_id, sample_body_site, visit_number) %>% 
  # Use metadata to select data collected at
  # (1) the same visit (across the two experiments)
  intersect_colData(by = c("subject_id", "sample_body_site",
                           "project_name", "study_full_name", "subject_gender",
                           "subject_race", "visit_number")) %>% 
  # and (2) the first visit
  filter_colData(visit_number == 1) %>% 
  # Remove OTUs not observed in any subject
  trim_empty_rows(phy16S) %>% 
  # For phy16S data: Convert taxa to proportions
  # Then, by row -- center (but do not scale) and normalize magnitude
  transmute(phy16S,
            cov = (t(counts)/colSums(counts))%>% 
              t() %>% 
              prep(., scale = FALSE)) %>% 
  # For cytokines data: by row -- center, 
  transmute(cyto,
            corr = prep(cyto_conc, scale = TRUE)) %>% 
  # Obtain list of assays (as matrices) from the MultiAssayExperiment
  assays() %>% 
  # Ensure that samples are in rows, variables are in columns
  lapply(., t) %>% 
  # Perform PCA on both experiments
  lapply(., function(x) {
    ade4::dudi.pca(x, scannf = FALSE, nf = 61, 
                   center = FALSE, scale = FALSE)
  })
```

We can see that we obtain the same results as in the workshop when performing the co-inertia analysis (available through `ade4`).

```{r}
coin <- ade4::coinertia(dudiX = pca_list[["phy16S"]], dudiY = pca_list[["cyto"]], 
                        scannf = FALSE, nf = 2)
coin$RV
```

### Results

```{r}
kable(head(coin$co),
      digits = 5) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r}
kable(head(coin$li),
      digits = 5) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r, echo=FALSE}
plot2 <- PlotCoinVars(coin, title = "TidyMultitude workflow",
             tab1 = "taxa", tab2 = "cytokines", 
             Labels1 = NULL,
             Labels2 = rownames(coin$li),
             label = TRUE,PtSize=2,LblSize=3,
             hjust = 0, vjust = -1.5)

par(mfrow = c(1, 2))
plot1; plot2
par(mfrow = c(1, 1))
```

